```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Does Poverty Predict Math Scores? Testing a School Board Member's Claim

### Introduction

A school board member in Montgomery County, Pennsylvania has claimed at a public meeting that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests. That's a testable claim, and as data journalists, our job is to test it.

We have school-level data from Montgomery County, Pa. that includes each school's FARMS percentage (the share of students eligible for Free and Reduced-price Meal System) and the percentage of students who met or exceeded math expectations on state tests. We'll use linear regression to evaluate whether the school board member's claim holds up, and we'll learn something important about what happens when you trust a model before examining your data.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_pa.csv")
```

### Task 1: Examine the Data Before You Plot Anything

Before we visualize anything, let's look at what we're working with.

```{r}
summary(schools$`FARMS Pct`)

schools |> arrange(desc(`FARMS Pct`)) |> head(10)

schools |> arrange(`FARMS Pct`) |> head(10)
```

Look at the distribution of FARMS percentages. Do any values look unusual or suspicious? What might it mean if a school has a FARMS percentage of exactly 100? Write two or three sentences describing what you see and whether anything concerns you.

ANSWER HERE
---
There are eight schools that have a FARMS percentage of 100, which appears suspicious to me. The possibilities are that these schools may overclaim that every student is eligible for FARMS, which may not be true, or there may be some errors with the data entry process. Another suspicious issue is the gap between schools with the highest FARMS percentage and those with the second-highest. It jumped from 100% to 53.9359, meaning that in this dataset, there are no schools with a FARMS percentage of roughly 54% - 99%, which seems very weird to me.   
---

### Task 2: Create a Scatterplot and Describe the Results

Now let's visualize the relationship between FARMS percentage and math scores for all schools.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County PA")
```

Describe what you see. Is there a clear trend? Does the regression line look like it represents the data well, or does it seem to be pulled in an odd direction? Connect what you observed in Task 1 to what you see here.

ANSWER HERE
---
There are no clear trends based on what I can see on this scatter plot. The regression line doesn't clearly represent the data points, even though it seems to be pulled a little bit upward, as all of the data points are spread out all over the place. There is a clear gap on the X-axis from roughly 54% to 99%, which resonates with my suspicion in the previous observation. 
---

### Task 3: Run the Regression Model

Let's build the linear regression model using all schools.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

There are three things to examine: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct.

The **p-value** tells you whether the relationship is statistically significant. The **Adjusted R-squared** tells you how much of the variation in math scores is explained by FARMS percentage — it's already a percentage, so a value of 0.025 means 2.5%, not 25%. The **coefficient** on FARMS Pct tells you the direction and size of the relationship: positive means higher FARMS predicts higher math scores; negative means higher FARMS predicts lower scores.

Report all three values and describe what they tell you. Does this model support the school board member's claim? Is this even a model worth using?

ANSWER HERE
---
(1) P-value is 0.1225 (higher than 0.05), meaning that this relationship is not statistically significant, or the relationship is very likely to happen just by chance. 
(2) Adjusted R-squared is 0.0254 or 2.54%, meaning that the FARMS percentage explains a very small portion (2.54%) of changes in math score.
(3) Coefficient is 0.12349, meaning that the FARMS percentage and math score change in the same direction, and the match score is likely to change 0.12349 point per one point change in FARMS percentage.  
---

### Task 4: What's Driving the Problem? Filter and Rerun.

In Task 1, you likely spotted schools with a FARMS percentage of 100. Those values are suspicious — in most districts, having *every* student on free/reduced lunch is unusual and may indicate a reporting artifact or data error. Let's see what happens when we exclude them.

```{r}
schools_filtered <- schools |> filter(`FARMS Pct` < 100)

model_filtered <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools_filtered)
summary(model_filtered)
```
Now compare the two models side by side:

```{r}
cat("Full dataset model:\n")
cat("  Coefficient:", coef(model)[2], "\n")
cat("  p-value:", summary(model)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model)$adj.r.squared, "\n\n")

cat("Filtered dataset model (FARMS < 100):\n")
cat("  Coefficient:", coef(model_filtered)[2], "\n")
cat("  p-value:", summary(model_filtered)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model_filtered)$adj.r.squared, "\n")
```

How did removing those schools change the model? Did the coefficient change direction? Did the p-value or R-squared improve meaningfully? What does this tell you about the influence of suspicious data values on a regression model?

ANSWER HERE
---
The coefficient is in the same direction, and the relationship size slightly increases from 0.12 to 0.40. The p-value improves meaningfully from 0.12 to 0.04, meaning that the second model illustrates that the relationship is unlikely to happen by just chance; nonetheless, the adjusted R-squared doesn't improve meaningfully as it changes from 2.54% to 6.48%, which is still a small portion of how the predictor can explain the changes in outcome. This tells us that suspicious or error values could make it more difficult for the model to determine a meaningful relationship between the predictor and outcome.     
---

### Task 5: Even the Better Model Has Problems

Even after filtering, the Adjusted R-squared remains very low. This means FARMS percentage — even when the data is clean — explains very little of the variation in math scores across these schools.

Look at the filtered scatterplot:

```{r}
schools_filtered |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage (Excluding 100% FARMS Schools)")
```

With a low R-squared, the residuals from this model aren't very meaningful as labels of "over-performing" or "under-performing" — the model doesn't know enough to make that judgment reliably. But let's calculate them anyway to see the spread:

```{r}
schools_filtered <- schools_filtered |>
  mutate(
    predicted = predict(model_filtered),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools_filtered |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

What are the largest residuals? Given the model's weak R-squared, how much confidence would you have in labeling these schools as over- or under-performers? Why?

ANSWER HERE
---
The largest residual is 36.66 at Gulph Elementary School. I wouldn't be very confident to label schools with large residuals over- or underperformed because as the R-squared is very low, meaning the FARMS percentage can explain a small portion in the variability of math score, and all data points spread out all over the place, there are too many data points that have large residual both above and below the line, so the regression line created by this model doesn't seem justify to be a standard. Hence, it will not be fair to call any schools below or above the line under- or overperformed, as the standard is not justifiable.   
---

### Task 6: Write the Story Pitch

You've now run two models, examined the data quality and seen that even a cleaned-up model explains very little of the variation in math scores. Return to the school board member's claim: that schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch (3-5 sentences) that addresses the following: Does the data support or contradict the claim? What data quality issues did you find? What would you need - different data, a different model, additional reporting - before you could responsibly publish a story about this claim? Who would you call first?

ANSWER HERE
---
The data we have is not strong enough to prove whether the claim is true, and despite the flaws in the data, it suggests that the claim is false, as many schools with high FARMS percentage perform very well in math tests. The quality issues we found are there are eight schools with 100% of students who are eligible for the FARMS, which is possibly caused by errors in the data entry process, and there are an unusual gap between school with most percentage of FARMS (100%), and the second highest (53.93%), meaning there are no schools with FARMS percentage roughly between 54% and 99%, which is very suspicion. Based on preliminary statistical analyses of this dataset, the relationship between FARMS percentage and math score is very likely due to chance (p-value = 0.12). Moreover, the FARMS percentage can only explain a small portion of changes in the math test (Adjusted R-squared = 2.54%). I will need a different dataset that excludes error values, a different model that could indicate a strong and meaningful relationship (p-value less than 0.05, high percentage of adjusted R-square), and an additional reporting task to retrieve more data from schools in other counties in Pennsylvania, in order to responsibly publish a story about this claim. I would talk to the school board of Montgomery County first to rectify the likely error data points, and then reach out to the school boards from other counties to collect more data.   
---
