---
title: "Polling Simulation: Sampling and the Central Limit Theorem"
---

In this activity, you'll use data from the 2024 Cooperative Election Study (CES) — a large-scale national survey conducted during the 2024 election — to simulate what happens when a news organization tries to estimate public opinion from a poll sample. The full dataset has nearly 55,000 respondents, which we'll treat as our population.

Your goal is to see firsthand how sample size affects the accuracy of estimates, and what the distribution of sample estimates actually looks like — the core idea behind the Central Limit Theorem.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from this URL and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

We'll focus on one question: what share of American adults identify as religiously unaffiliated (atheist, agnostic, or nothing in particular)? This has been a major story in American religion journalism for years.

First, we'll create a new column called `unaffiliated` that is `TRUE` if a respondent's `trad` value is `"Atheist"`, `"Agnostic"`, or `"Nothing in Particular"`, and `FALSE` otherwise. Then calculate the proportion of the full dataset that is unaffiliated. Save this value as `population_proportion`.

```{r}
ces <- ces |>
  mutate(unaffiliated = trad %in% c("Atheist", "Agnostic", "Nothing in Particular"))

proportion <- mean(ces$unaffiliated)
```

That population proportion is the "true" population value that any poll is trying to estimate. Write it down — you'll be comparing everything else to it.
---
Proportion = 35.99%
---
### Task 4: Take a single small sample

Use `set.seed(42)` and then sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of unaffiliated respondents in your sample.

```{r}
set.seed(42)

sample_50 <- ces |>
  sample_n(50)

proportion_50 <- mean(sample_50$unaffiliated)
proportion_50
```

How close is your sample estimate to the original proportion? Is it surprising?

ANSWER HERE
---
Pretty close. The proportion of the unaffiliated in my 50 samples is 32%, which is not that surprising, as it's quite close to 35.99%
---
### Task 5: Take a single larger sample

Now sample 500 rows. Set a different seed value, grab the sample and save it as `sample_500`. Calculate the proportion of unaffiliated respondents in this sample.

```{r}
set.seed(42)

sample_500 <- ces |>
  sample_n(500)

proportion_500 <- mean(sample_500$unaffiliated)
proportion_500
```

How did the estimate change compared to the sample of 50? What does this suggest about sample size?

ANSWER HERE
---
It's closer. The proportion of my 500 samples is 36.2%, compared to that of 50 samples, which is 32%; it's a lot closer to 35.99%. This could suggest that the bigger the sample size, the closer the representation is to the whole population.
---
### Task 6: Simulate many polls

A single sample can get lucky or unlucky. To understand how polling actually works, we need to see the *distribution* of estimates across many samples. The code below draws 1,000 samples of size 50, calculates the proportion of unaffiliated respondents in each, and stores the results.

```{r}
set.seed(5555555)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$unaffiliated)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side — one for the 1,000 samples of size 50 and one for the 1,000 samples of size 500. Add a vertical line at `proportion` to each so you can see how well the samples are centering on the true value.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share Religiously Unaffiliated",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()
```

### Task 8: Reflect on what you see

**What shape do both histograms have?** Describe them in plain language. Does it surprise you that this shape emerges from a variable that is simply TRUE or FALSE for each respondent?

ANSWER HERE
---
It looks quite similar to the bell shape, with most of the data accumulated at the center (mean) and almost symmetric tails on both sides. This does not surprise me that much because I watched a clip you previously assigned to do last week that explain about the Central Limit Theorem, but the first time I learned about this concept, it was pretty intriguing because intuitively I would think that the distribution results, regardless of how many times you sample it, would be in random shape not close to the bell-shape.   
---
**Compare the two histograms.** What is the most visible difference between the n=50 and n=500 distributions? What does that difference tell you about what happens when a poll uses a larger sample?

ANSWER HERE
---
The base of the shape of the n50 distribution is wider than that of n500, or in another way, the data seemed to be more clustered around the center (the mean of the true population) for n500 compared to that of n50. This could mean that when a poll uses a larger sample, the closer to the real representation of the population it will get.
---
**Both distributions are centered near the red line.** What does it mean that 1,000 different random samples — each one drawn independently — tend to cluster around the same value? What would it mean for journalism if that *weren't* the case?

ANSWER HERE
---
This could mean that the number of my sample is large enough, so the distribution of results I get looks close to a bell-shaped curve, or it somewhat aligns with the Central Limit Theorem. But if this were not the case, for journalism, that means this sample size is too small, so no matter how many times you run the sampling, the results tend to jump around, so the distribution of them will not tend to cluster around the same value.   
---
**A reporter tells you their story is based on a poll of 50 people.** Based on what you've seen in the simulation, what questions would you want to ask before trusting that number? You don't need to use any technical terms — just describe what you observed.

ANSWER HERE
---
I would ask what the size of the population these 50 people are drawn from. If, compared to the population size, the sample of 50 people seems to be too small, I might not trust that this number could represent the real population.   
---

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
