---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80
```{r}
set.seed(42)

population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

population
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sampled_data <- population |>
  slice_sample(n = 100)

sampled_data
```


### Task 4: Calculate the mean age and standard deviation of the sample
```{r}
sample_stats <- sampled_data |>
  summarise(
    mean_age = mean(age),
    sd_age = sd(age)
    )

sample_stats
```


### Task 5: Calculate the mean age and standard deviation of the population
```{r}
population_stats <- population |>
  summarise(
    mean_age = mean(age),
    sd_age = sd(age)
  )

population_stats
```


### Task 6: Compare the sample and population statistics
```{r}
sample_stats <- sample_stats |>
  mutate(data = "sample") |>
  select(data, mean_age, sd_age)

population_stats <- population_stats |>
  mutate(data = "population") |>
  select(data, mean_age, sd_age)

comparison <- sample_stats |> bind_rows(population_stats)
comparison
```


Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

```{r}
smaller_sampled_data <- population |>
  slice_sample(n = 20)

smaller_sample_stats <- smaller_sampled_data |>
  summarise(
    mean_age = mean(age),
    sd_age = sd(age)
    )

smaller_sample_stats <- smaller_sample_stats |>
  mutate(data = "smaller sample") |>
  select(data, mean_age, sd_age)

comparison <- comparison |> bind_rows(smaller_sample_stats)
comparison
```

---
With a smaller sample, the means I got are quite similar (population = 49.87, sample = 48.77, smaller sample = 49.95), but the SD is larger with the smaller sample (population = 17.51, sample = 18.91, smaller sample = 19.20) 
---
When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
