---
title: "SoccerFouls"
---

Do fouls matter in women's college soccer? In this assignment, you'll be working with a dataset that contains information about soccer matches from last season. You'll load the data, perform some basic statistical analysis, and create a visualization to explore the relationship between fouls and the goal differential (and thus the outcome). Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the soccer dataset from https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv

```{r}
matches <- read_csv("https://raw.githubusercontent.com/dwillis/NCAAWomensSoccerData/refs/heads/main/data/ncaa_womens_soccer_matchstats_2024.csv")
```


### Task 3: Create a new column called `goal_diff` that calculates the difference between the two teams' goals.

```{r}
matches <- matches |> mutate(goal_diff = team_score - opponent_score)
```


### Task 4: Use `summarize()` and the `cor()` function to calculate the correlation between fouls and goal differential, using the pearson method.

```{r}
matches |> summarize(correlation = cor(fouls, goal_diff, method="pearson"))
```

### Task 5: Create a scatterplot of fouls and goal differential, using `geom_point()`

```{r}
matches |> 
  ggplot() +
  geom_point(aes(x=goal_diff, y=fouls)) +
  geom_smooth(aes(x=goal_diff, y=fouls), method="lm")
```

### Task 6: Describe the results of both the correlation calculation and the scatterplot.
---
The correlation result of 0.018 and the nearly flat trend line on the scatter plot indicate that fouls and goals differ are barely correlated. Simply speaking, the number of fouls is very unlikely to determine whether a game is won or lost.   
---
### Task 7: Choose another variable (rec: sh_att/so_g) and repeat the steps in the steps in Task 4 - 5, then describe the result

```{r}
matches |> summarize(correlation = cor(sh_att, goal_diff, method="pearson"))
```

```{r}
matches |> 
  ggplot() +
  geom_point(aes(x=goal_diff, y=sh_att)) +
  geom_smooth(aes(x=goal_diff, y=sh_att), method="lm")
```
---
The correlation result is 0.56, and even though it's not 45 degrees to the x-axis, the trend line on the scatter plot is still sloped. These indicate that shot attempt and goal differences somewhat have a correlation. The shot attempt is one of the factors that affect the game result, but it is not a dominant factor. 
---
When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
