---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Molecule Jongwilai"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library("tidyverse")
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
restaurant_health_inspection <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- restaurant_health_inspection |>
  summarise(
    mean_compliance = mean(compliance_score),
    sd_compliance = sd(compliance_score),
    min_compliance = min(compliance_score),
    max_compliance = max(compliance_score)
  )
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
restaurant_health_inspection |>
  ggplot() +
  geom_histogram(aes(x = compliance_score), binwidth = 4) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +
  labs(title = "The majority of restaurants scored high at 100 for compliance score",
       x = "Compliance Score",
       y = "Number of Restaurant")
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**
---
The standard deviation of the scores is just nearly 5.83, meaning the majority of the scores are clustered around the mean of 96.30 and rarely spread out. The histogram tells us that this dataset doesn't distribute normally and is heavily skewed to the left, meaning that most of the restaurants get high scores, while a small group of restaurants get lower scores. The interesting aspects here are that most of the restaurants in Montgomery County are really healthy (may have to recheck what "health inspection" is), or the inspectors just go easy on them? And what are the restaurants that get a lower score? What makes them different from most of the restaurants? My next step would be to take a closer look at the criteria or how the inspectors score these restaurants, and go look into restaurants that get high scores and low scores to see whether those scores could actually reflect the health conditions at the restaurants.   
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
ridership <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_status <- ridership |>
  summarise(
    mean_bus = mean(bus),
    sd_bus = sd(bus),
    mean_rail = mean(rail),
    sd_rail = sd(rail)
  )
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")

mean_bus <- mean(wmata_daily$bus)
sd_bus <- sd(wmata_daily$bus)
mean_rail <- mean(wmata_daily$rail)
sd_rail <- sd(wmata_daily$rail)

wmata_daily <- wmata_daily |>
  mutate(rail_zscore = (rail - mean_rail)/sd_rail) |>
  mutate(bus_zscore = (bus - mean_bus)/sd_bus)
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_daily |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore))) |>
  select(date, weekday, rail, rail_zscore)
```

ANSWER
---
There are 6 unusual days with the low use of rails. 
---

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(42)

wmata_daily_91 <- wmata_daily |>
  slice_sample(n = 91)

wmata_daily_91 |>
  summarise(
    mean_bus = mean(bus),
    sd_bus = sd(bus),
    mean_rail = mean(rail),
    sd_rail = sd(rail)
  )
```
---
For the bus, the mean and SD of the sample of 91 days are slightly higher than those of the population, with a difference of 3215 for the mean and 6442.29 for the SD. Similarly, for rail, the mean and SD of the sample of 9 days are also a little bit higher compared to those of the population, with a 6306 difference for the mean and 8394.4 in SD. This could suggest that 91 days is a large enough sample size, as its mean and SD are quite close to those of the real population.   
---

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)
---
The z-score analysis could indicate days with unusually low levels of rail ridership, prompting questions about what occurred on those days that led people to commute by rail less than on other days. There might be accidents or major maintenance during those days that could be a story.  
---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
car_thef <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- car_thef |>
  mutate(theft_rate = (thefts_2023/population)*10000)
```

---
For a meaningful comparison, the same base is required. The raw number is not useful for the comparison because each county has a different population. The counties with more population may have more raw numbers of thefts, but may not make their situation more severe than the counties with less population that may have fewer raw numbers of thefts because their rate may be higher. Hence, the per-capita rate is more useful for meaningful comparison as it calibrates every raw count, based on the same base.  
---

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts |>
  summarise(
    med_theft = median(theft_rate),
    count_theft = sum(thefts_2023)
  )

md_car_thefts |>
  filter(theft_rate > 9.58)
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

---
The calculation reveals that half of all counties (12 out of 24) in Maryland have a high rate of car thefts compared to the median. This could suggest that car theft seems to be quite serious issue in Maryland, and this could be the lede for the story that may be investigate furthered in, like why the car theft rate is so high or is the law enforcement in Maryland weak when it comes to crime about cars?  

---
## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)
---
I would use Z-Scores to find the pools that may underperform, as the SD is not too high, that mean most of the observations are clustered near the mean at the center, so Z-Scores that are higher or lower than 2 could be rare, which could be potential outliers. 
---

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)
---
I would find out the pools that have Z-Scores lower than -2, which could suggest the unusually low level of average use compared to those of other pools. Z-scores are most useful becuase it compares the performance of each data point to its peers, based on the same standardized base (mean). It would be hard to compare performance by raw numbers, as it doesn't have a central base to compare to. 
---

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

---
I would ask whether this sample is drawn randomly and find out what the real total count is to see if the sample is too small or not. If the sample size is too small, it may not be able to accurately represent the real population, and that could raise a concern in reporting any information from this sample.
---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
