---
title: "HW10: Significance Testing & Effect Size"
name: Molecule Jongwilai
---

You previously calculated the mean, margin of error and confidence interval for monthly car thefts in Prince George's County. Your editor now wants to know: it _looks_ like car thefts are decreasing lately, but is that change real, and how big is it? In this assignment, you'll use significance testing and effect size to answer those questions. Load the tidyverse and the data to get started.

### Task 1: Load the tidyverse and the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

```

### Task 2: Split the data and test for significance (4 points)

We'll split the monthly data into two halves, and then ask whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05. First, fill in the hypotheses, then replace the REPLACE_ME values in the code.

**Null Hypothesis:** The car thefts have not decreased in the second half.

**Alternative Hypothesis:** The car thefts have actually decreased in the second half.

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |>
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

```

**Answer this question:** What is the p-value? Is it less than 0.05? Can we reject the null hypothesis? What does this tell your editor?

PUT ANSWERS HERE
---
The p-value is 0.0000214. It is less than 0.05, meaning that we can reject the null hypothesis. We can tell the editor that the car thefts are very likely to decrease in the second half, and not by chance (the possibility that the car thefts will not change is very unlikely). 
---

### Task 3: Visualize the comparison (4 points)

Let's see the two periods side by side with their confidence intervals. Replace the REPLACE_ME values.

```{r}
# Summarize by period
period_summary <- car_thefts |>
  group_by(period) |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts,
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |>
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 5)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Answer this question:** Look at the error bars (confidence intervals) on the chart. Do they overlap? What does that tell you about whether the difference between the two periods is real?

PUT ANSWERS HERE
---
The confidence intervals of both halves do not overlap, meaning that we're very confident that the true average of both halves, which will fall within those aforementioned ranges, will not be the same. This also means that it's very likely the difference between the two periods is real, as the true mean of the second half will be less than that of the first half.  
---

### Task 4: Calculate the effect size (5 points)

We've established that the decline is statistically significant. But how large is it? Let's calculate Cohen's d to measure the effect size. For a two-sample comparison, we use the pooled standard deviation:

**Pooled SD** = sqrt(((n1 - 1) × sd1² + (n2 - 1) × sd2²) / (n1 + n2 - 2))

**Cohen's d** = (mean1 - mean2) / pooled SD

Replace the REPLACE_ME values. You'll need to pull the values from `period_summary`.

```{r}
# Extract the values from period_summary
first_half_stats <- period_summary |> filter(period == "First Half")
second_half_stats <- period_summary |> filter(period == "Second Half")

# Calculate pooled standard deviation
pooled_sd <- sqrt(
  ((first_half_stats$months - 1) * first_half_stats$sd_thefts^2 +
   (second_half_stats$months - 1) * second_half_stats$sd_thefts^2) /
  (first_half_stats$months + second_half_stats$months - 2)
)

# Calculate Cohen's d
cohens_d <- (first_half_stats$mean_thefts - second_half_stats$mean_thefts) / pooled_sd

# Interpret it
effect_label <- case_when(
  abs(cohens_d) < 0.2 ~ "Negligible",
  abs(cohens_d) < 0.5 ~ "Small",
  abs(cohens_d) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `First half mean` = round(first_half_stats$mean_thefts, 1),
  `Second half mean` = round(second_half_stats$mean_thefts, 1),
  `Difference` = round(first_half_stats$mean_thefts - second_half_stats$mean_thefts, 1),
  `Cohen's d` = round(cohens_d, 3),
  `Interpretation` = effect_label
)
```

**Answer this question:** How would you describe this effect size? Compare it to the effect sizes we calculated in class (the response times activity had a small effect, the housing prices activity had a medium effect). What does a large effect size mean for the story you'd write about car thefts?

PUT ANSWERS HERE
---
The effect size is large, meaning that the change in car thefts is more meaningful statistically compared to the changes in response times and house prices from the class activity. Let's say if there is a policy or practice that could lead to this level of decrease in car thefts, it's highly worth implementing. A large effect size means that we can report that this change is highly important statistically.  
---

### Task 5: Putting it all together (7 points)

You now have three pieces of evidence: the confidence intervals from last week, the significance test from this assignment, and the effect size.

1. In Part 1, your editor asked "just tell me if car thefts are going up or down". Now you have significance testing. Write 2-3 sentences telling your editor the answer, using all three pieces of evidence (CI, p-value, and effect size).

PUT ANSWERS HERE
---
From July 2023 to Feb 2026, car thefts in Prince George's are very likely to go down. Splitting the period to 2 halves, we found that the mean of the first half is nearly 439 incidents per month, plus or minus 41, and the mean of the second half is about 312, plus or minus about 34. In addition, the p-value of the decrease is less than 0.05, meaning that the change is likely to be true or doesn't happen by chance, and the effect size of 1.791 tells us that this change is very meaningful statistically.    
---

2. Write a headline and a 2-3 sentence lede for a story about car thefts in Prince George's County. Unlike the response times story from class (where the effect was small and the headline needed to be cautious), this is a case where the data strongly supports a clear conclusion. Your writing should reflect that — but still note the limitations (correlation vs. causation, what else might explain the decline).

PUT ANSWERS HERE
---
Headline: Prince George's found significant decline in car thefts

Lede: Data shows that the car thefts between November 2024 and February 2026 in Prince George's significantly declined at the average of about 312 incidents per month, 127 cases lower than that of the previous 16 months. Despite the clear, significant drop, multiple factors could account for this change.   
---

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
