---
title: "Multiple Regression Analysis: Maryland Voter Turnout"
name: "Molecule Jongwilai"
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships in civic data, such as how different demographic and socioeconomic factors collectively impact voter turnout. By understanding these relationships, journalists can provide deeper insights into civic participation patterns and identify potential barriers to voting.

```{r setup}
library(tidyverse)
#install.packages('corrplot')
library(corrplot)
```

## Load the Data

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains information about voter turnout and various demographic factors across Maryland counties. This dataset will be used to build our model that will try to see which factors are most important in predicting voter turnout. The turnout percentage is from the 2024 general election.

```{r}
md_turnout <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_turnout.csv")
head(md_turnout)
```

## Task 1: Exploratory Data Analysis (5 points)

Before building our model, let's explore the data to understand the variables and their relationships. Examine the correlation matrix to see how the independent variables relate to each other and to the dependent variable (voter turnout). Which variables seem to be correlated? Are there any that are not correlated at all? Also look at the multiple scatterplots. Which variables seem to have a linear relationship with the dependent variable? Which do not?

ANSWER HERE
---
For the voter turnout, percentage of the unemployed, percentage of white, and median age seem to be variables that have some degree of correlation with it (-0.74, 0.69, and 0.50), while population, income, and percentage of higher education don't show a strong correlation with voter turnout (-0.39, 0.45, 0.30).
...
For relationships between other variables;
Variables that seem to be correlated to each other to some degree (abs(correlation) > 0.5) 
- population and percentage of higher education (0.54)
- population and percentage of white (-0.67)
- income and percentage of higher education (0.75)
- percentage of the unemployed and percentage of white (-0.59)
- median age and percentage of white (0.53)
Variables that don't show a strong correlation or don't seem correlated at all 
- population and income (0.34)
- population and percentage of the unemployed (0.41)
- population and median age (-0.40)
- income and percentage of the unemployed (-0.32)
- income and media age (-0.21)
- income and percentage of white (-0.12)
- percentage of higher education and percentage of the unemployed (-0.22)
- percentage of higher education and median age (0.04)
- percentage of higher education and percentage of white (-0.22)
- percentage of the unemployed and median age (-0.3)
...
Examining from the scatterplots, I think percentage of the unemployed and percentage of white show quite clear linear relationship with voter turnout, as the slope is distinctly steep, and all data points seem to distribute consistently and closely around the line with small variations, while variable like population doesn't show a strong clear linear relationship with voter turnout, as the line is almost flat, and the data points seem to cluster on one end of the line but spread further on another end.   
---


```{r}
# Summary statistics
summary(md_turnout)

# Create correlation matrix
selected_vars <- md_turnout |> 
  select(turnout, population, income, pct_bachelors_higher, 
         pct_unemployed, median_age, pct_white)
correlation_matrix <- cor(selected_vars)
print(correlation_matrix)

# Visualize correlation matrix
corrplot(correlation_matrix, method = "circle")

# Create scatterplots
md_turnout |>
  pivot_longer(cols = c(population, income, pct_bachelors_higher, 
                       pct_unemployed, median_age, pct_white),
               names_to = "variable", 
               values_to = "value") |>
  ggplot(aes(x = value, y = turnout)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(title = "Relationship between predictors and voter turnout",
       y = "Voter turnout (%)")
```

## Task 2: Initial Model Evaluation (5 points)

Next, we construct an initial linear model to analyze how different factors influence voter turnout across Maryland counties. This model helps us quantify the relationships and assess which factors are significant predictors of turnout. 

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For this model:

* **Min and Max** values indicate the range of residuals. 
* **Median** value tells us if the model is generally centered around the actual data points.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

* **Population**: Is the coefficient positive or negative? Is it statistically significant?
---
The coefficient is positive, but it isn't statistically significant (P-value > 0.05)
---
* **Income**: Does higher income correlate with higher turnout?
---
Yes, because the coefficient is positive.
---
* **Education (pct_bachelors_higher)**: How does education level impact voter participation?
---
The people with higher education seem to participate less in the voting, but this correlation isn't statistically significant.
---
* **Unemployment**: Is there a relationship between unemployment and voter turnout?
---
Yes, they have a negative coefficient, but the relationship isn't statistically significant.
---
* **Median Age**: Do counties with older populations have different turnout rates?
---
Yes, the older age seems to impact the higher turnout, as the coefficient is positive.
---
* **Race (pct_white)**: Is there a relationship between racial composition and turnout?
---
Yes, the higher percentage of white seem to affect the higher turnout rate.
---

### Model Fit

* **Residual Standard Error**: How much does the model typically deviate from actual turnout?
---
0.02174, which is very small.
---
* **Multiple R-squared**: What percentage of the variability in turnout is explained by the model?
---
85.39%
---
* **Adjusted R-squared**: How well does the model fit after accounting for the number of variables?
---
80.23%, so I would say the model fit quite well.
---
* **F-statistic**: Is the model as a whole statistically significant?
---
Yes, as the p-value is very low. 
---


Examine the output of the model, checking the coefficients, residuals, and overall model fit. How well does this multiple variable model predict turnout? Do any of the variables seem to be insignificant? Is this a useful model, based on what you've seen? Why?

ANSWER HERE
---
I think this model predict he turnout rate quite well, judging from the adjusted R-squared, which is 80.23%, meaning that these multiple variables can explain a large portion (80.23%) of the variability in the turnout. However, I think variables like population, higher education, and unemployment seem to be insignificant as their P-values are high (more than 0.05). Overall, I think this model is useful as it shows statistical significance between predictors and outcomes (very low P-value), and it could explain a big portion of the changes in outcome indicated by a high R-squared. 
---

```{r}
# Build initial model with all predictors
initial_model <- lm(turnout ~ population + income + pct_bachelors_higher + 
                   pct_unemployed + median_age + pct_white, data = md_turnout)
summary(initial_model)

```

## Task 3: Model Refinement (5 points)

Create three alternative models by adding or removing variables to see if we can improve the model's performance. Turnout will always be the dependent variable, but you should alter the independent variables by removing or adding some. Then compare the models using adjusted R-squared values. Are any of the models significantly better than the others?

ANSWER HERE
---
Based on my experiment, model 3, which I include every variables except population and education, has the most adjusted R-squared at 82.01%, which is better than model 2 (every variables except education), model 4 (every variables except population, education, and unemployment), and initial model (every variables) that have adjusted R-squared at 81.28%, 80.24%, and 80.23% respectively. 
---

```{r}
# Model 2
model2 <- lm(turnout ~  population + income + pct_unemployed + median_age + pct_white, data = md_turnout)
summary(model2)

# Model 3
model3 <- lm(turnout ~ income + pct_unemployed + median_age + pct_white, data = md_turnout)
summary(model3)

# Model 4
model4 <- lm(turnout ~ income + median_age + pct_white, data = md_turnout)
summary(model4)

# Compare models
cat("Model 1 (Full model) - Adjusted R-squared:", summary(initial_model)$adj.r.squared, "\n")
cat("Model 2 - Adjusted R-squared:", summary(model2)$adj.r.squared, "\n")
cat("Model 3 - Adjusted R-squared:", summary(model3)$adj.r.squared, "\n")
cat("Model 4 - Adjusted R-squared:", summary(model4)$adj.r.squared, "\n")

# Create comparison chart
model_names <- c("Full model", "Model 2", "Model 3", "Model 4")
adj_r_squared <- c(summary(initial_model)$adj.r.squared,
                   summary(model2)$adj.r.squared,
                   summary(model3)$adj.r.squared,
                   summary(model4)$adj.r.squared)

model_comparison <- data.frame(Model = model_names, Adjusted_R_squared = adj_r_squared)
ggplot(model_comparison, aes(x = reorder(Model, Adjusted_R_squared), y = Adjusted_R_squared)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison by Adjusted R-squared",
       x = "Model")
```


## Task 4: Using the Model (3 points)

Without using the words "multiple regression model", describe how you could use this analysis to cover an election or voter participation story.

ANSWER HERE
---
I could use this model to evaluate which variable has the most effect on higher or lower vote turnout. This model could be very useful for identifying which variables are or are not influential to voter participation, as the model compares all variables collectively, which could be more methodical than looking into each of them in isolation. Furthermore, I can use this analysis to explain why some counties have high or low turnout rates by highlighting the disparities in the variables that seem to affect the turnout rate most in each county. This information could also be useful in framing questions for interviews with experts to elaborate on the implications of this data.  
---
