---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_song <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_song |>
  summarise(
    mean = mean(danceability),
    sd = sd(danceability)
  )
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_song |>
  ggplot() +
  geom_histogram(aes(x = danceability))
```

### Task 5: Add a line to the histogram for the mean danceability score

Write code to add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_song |>
  ggplot() +
  geom_histogram(aes(x = danceability)) +
  geom_vline(aes(xintercept = mean(danceability)), color = "purple", linetype = "dashed", size = 1)
```

### Task 6: Describe our results

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and distribution tell us about the songs in the dataset?
---
This dataset does not distribute normally, as the bell-shaped curve is skewed to the left. The mean of this dataset tells us that most of the song have have a "danceability" score clustered around 0.67, or two-thirds of all songs fall between scores 0.51 and 0.83 regarding the SD value, meaning most of the songs (as I understand this score) are quite "danceable". And as this is not a standard distribution, the average may not be the proper measure to represent this dataset.    
---
### Task 7: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

