---
title: "Final Project"
author: "Molecule Jongwilai"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

**4/19/2026**

The main hypotheses; 
- Could differences in kids' math scores be explained by countries or regions?
- Which variables could be associated with the kids' math proficiency?

Based on my key questions and data I acquired, I think the following analyses would be the most applicable and useful to test my hypotheses. 

*Percentile*
- I may begin by calculating the percentile of each math score to provide some sense of hierarchical locations for the references in the further steps.  

*Normal Distribution / Z-score*
- I want to create a histogram to check the distribution of overall math scores. If the distribution shows a large SD or a far spread out of the observations from the mean, I could calculate the Z-score of each observation to identify obvious outliers and examine which students from which countries the usually high scores belong to.     
- I also want to check the distribution of other datasets of factors—socioeconomic index, study time, and resources—to see if they are distributed normally, and if they display large SD. If there are significant outliers, I can look into more on what scores those outlier data points represent, and what the locations (percentiles) of those scores are, to see if the abnormality of these factors is associated with higher scores.     

*Standard Error / Confidence Interval*
- I want to compare the average math score of kids from each region, visualizing bar charts or box plots, but for a valid comparison that could capture the nuance of uncertainty, I also want to calculate the CI of each average and factor them into the analysis.   

*Correlation / Multiple Linear Regression* 
- I want to prove whether any additional variables, besides regions, could potentially influence the math scores. I want to calculate correlation and plot a scatter plot that illustrates the relationship between math scores and socioeconomic index, study time, and educational resources. And if you will cover this in class, I may also use multiple regression to assess the combined effect of these variables. 

*Statistical Significance / Effect Size*
- Lastly, if time allows, I want to assess the statistical significance to determine, for example, whether the difference in math score from Asia differs from the overall mean by chance, or whether the relationship between math score and study time occurs due to chance. Also, I will consider the effect size to evaluate the significance of the difference between regions or relationships between variables.      

----------------------------------------------------------------------------------------------

**4/11/2026**

I downloaded the PISA 2022 student questionnaires data from this site: https://webfs.oecd.org/pisa2022/index.html, which includes the math scores of 613,744 students and other variables I need for the analysis, such as country, socioeconomic status, educational resources, etc. 

The original file is frighteningly gigantic (like 2GB), so I had to preliminary clean it up and select only the variables I want to work with. 

I used the codebook and the master version of the questionnaire (I put them in the final_project folder as well) to interpret and designate variables I want to analyze as follows;

1. CNT: Country Code
2. Math Score: PV1MATH (Plausible value 1 in mathematics)
3. Socioeconomic Status: ESCS (Index of economic, social, and cultural status)
4. Study Time:
4.1 ST294Q02JA (How many days in a typical school week does the student study for school or homework before go to school?)
4.2 ST295Q02JA (How many days in a typical school week does the student study for school or homework before leaving school?)
5. Educational Resources:
5.1 ST255Q01JA (How many books are there in the student's house?)
5.2 ST253Q01JA (How many digital devices with screens are there in a student's house?)
5.3 ST254Q06JA (How many cell phones with internet access are there in the student's house?)

After targeting variables, I brought in the PISA 2022 student questionnaires data to an R notebook, selected said variables, and saved a new file as .csv.

```{r}
#I did this elsewhere because the file is really big but want to note the process here in case you want to check my work.

#library(tidyverse)
#install.packages("haven")
#library(haven)
#student <- read_sav("student.sav")
#library(dplyr)
#student_data_subset <- student |> select(CNT, PV1MATH, ESCS, ST294Q02JA, ST295Q02JA, ST255Q01JA, ST253Q01JA, ST254Q06JA)
#write.csv(student_data_subset, "student_data_subset.csv", row.names = FALSE)
```

I brought the new data file here and cleaned it, so the column names sound understandable.

```{r}
math_score <- read_csv("student_data_subset.csv")
math_score <- math_score |>
  rename(
    country = CNT,
    math_score = PV1MATH,
    socioecon_index = ESCS,
    studytime_before_school = ST294Q02JA,
    studytime_after_school = ST295Q02JA,
    books_count = ST255Q01JA,
    digital_device_count = ST253Q01JA,
    phone_internet_count = ST254Q06JA
  )
glimpse(math_score)
```

Note about the dataframe; 

There are 613744 entries representing 613744 students PISA surveyed in 2022. 
There are 8 variables 

1. country: The values are still in country code format. I'll clean them in the next step.

2. math_score: The values range from 0 to 943.041

3. socioecon_index: The values range from -6.8407 to 7.3800. I'll read this site (https://www.oecd.org/en/publications/pisa-2022-results-volume-v_c2e44201-en/full-report/component-20.html?utm_source=chatgpt.com) to get a grasp on how the index indicates socioeconomic status in the next step. 

4. studytime_before_school: The values need to be converted into these labels. 
01 = 0 days
02 = 1 day
03 = 2 days
04 = 3 days
05 = 4 days
06 = 5 or more days

5. studytime_after_school: The values need to be converted into these labels.
01 = 0 days
02 = 1 day
03 = 2 days
04 = 3 days
05 = 4 days
06 = 5 or more days

6. books_count: The values need to be converted into these labels.
01 = no books
02 = 1 - 10 books
03 = 11 - 25 books
04 = 26 - 100 books
05 = 101 - 200 books
06 = 201 = 500 books
07 = more than 500 books

7. digital_device_count: The values need to be converted into these labels;
01 = no devices
02 = 1 device
03 = 2 devices
04 = 3 devices
05 = 4 devices
06 = 5 devices
07 = 6 - 10 devices
08 = more than 10 devices

8. phone_internet_count: The values need to be converted into these labels;
01 = no cell phones with internet 
02 = 1 or 2 cell phones with internet
03 = 3 - 5 cell phones with internet
04 = more than 5 cell phones with internet
05 = don't know

If time allows, I might try to analyze other relevant variables, or if you think other variables may correlate to the match score and want me to include them in the project, please let me know. Thank you.
